Cellular Automata and Computational Universality

### 2018-08-26 paper/paper.tex
Paper updates, added more life-like rules

    Reading the bottom row of this transition table, we get the binary number
    $01011010_{\text{bin}} = 90_{\text{dec}}$, allowing us to name this rule set
    Rule 90. Using this convention, we may now construct our own automata based
    solely on the knowledge of its name (at least for elementary cellular
    automata). For example, take Rule 110, introduced by Wolfram 1983
    \cite{WolframRule110}. We first convert $110_{10}$ to binary, and then use
    this to fill in the table as above:

@@

    \subsection{Other types of automata}
    \subsubsection{Reversible cellular automata}
    A cellular automaton is reversible if, for every configuration of the
    automaton, exactly one configuration that causes it (its preimage). If we
    consider the automaton as a function that maps configurations to
    configurations, then for a reversible automaton this function would be
    bijective. Where this is not the case, where not every configuration has a
    preimage, these configurations without preimages are called Garden of Eden
    patterns \cite{GardenOfEden}.
    \subsubsection{Totalistic cellular automata}
    \subsubsection{Proabilistic cellular automata}
    \subsubsection{Cyclic cellular automata}

\section{Computational Universality}

>>>>

    Reading the bottom row of this transition table, we get the binary number
    $01011010_{\text{bin}} = 90_{\text{dec}}$, allowing us to name this rule set
    Rule 90. Using this convention, we may now construct our own automata based
    solely on the knowledge of its name (at least for elementary cellular
    automata). For example, take Rule 110, introduced by Wolfram 1983
    \cite{WolframRule110}. We first convert $110_{\text{dec}}$ to binary, and
    then use this to fill in the table as above:

@@

    \subsection{Other types of automata}
    \subsubsection{Reversible cellular automata}
    A cellular automaton is reversible if, for every configuration of the
    automaton, exactly one configuration that causes it (its preimage). If we
    consider the automaton as a function that maps configurations to
    configurations, then for a reversible automaton this function would be
    bijective. Where this is not the case, where not every configuration has a
    preimage, these configurations without preimages are called Garden of Eden
    patterns \cite{GardenOfEden}.

    There are algorithms to compute whether or not a one-dimensional cellular
    automaton is reversible \cite{oneDReversible}; the same is not true for
    higher dimensions however, and computing reversibility for cellular automata
    in two-dimensions and up is undecidable \cite{twoDUndecidable}.

    \subsubsection{Totalistic cellular automata}
    A cellular automaton is totalistic if the value of cell $a$ at time $t$
    depends on the sum of the values of the cells in its neighbourhood at time
    $t-1$, with the values of cells being taken from some finite set
    \cite{totalistic}. An automaton is outer totalistic if it takes into account
    itself as well as the total of its neighbours at time $t-1$; Conway's Game
    of Life is an example of an outer totalistic cellular automaton where all
    cell values are 0 or 1, as are all other Life-like automata.

    \subsubsection{Probabilistic cellular automata}
    \subsubsection{Cyclic cellular automata}

\section{Computational Universality}
    A system of rules, be they from a computer's instruction set, programming
    language, or cellular automaton, is called computationally universal, or
    Turing-complete, if it can be used to simulate a Turing machine.

    \subsection{Turing Machines}
    A Turing machine is a mathematical model of computation which was first
    proposed by Alan Turing in 1936 \cite{SipserToC}. It consists of a infinite
    tape divided into cells each of which may contain one character from some
    finite alphabet, and a printer head capable of reading from and writing to the
    tape on the cell at which it is currently positioned. In a more formal
    sense, it is the seven-tuple $M=(Q, \Sigma, \Gamma, \delta, q_0,
    q_{\text{accept}}, q_{\text{reject}})$, where $Q, \Sigma, \Gamma$ are all
    finite sets and

    \begin{itemize}
        \item $Q$ is the set of states
        \item $\Sigma$ is the input alphabet not containing the blank symbol,
            $\textvisiblespace$
        \item $\Gamma$ is the tape alphabet where $\textvisiblespace \in \Gamma$
            and $\Sigma \subseteq \Gamma$
        \item $\delta: Q \times \Gamma \rightarrow Q \times \Gamma \times
            \{L,R\}$ is the transition function
        \item $q_0 \in Q$ is the start state
        \item $q_{\text{accept}} \in Q$ is the accept state, and
        \item $q_{\text{reject}} \in Q$ is the reject state, with
            $q_{\text{accept}} \neq q_{\text{reject}}$
    \end{itemize}
